---
title: 'EDA: COVID-19 and Brent Crude Oil'
author: "Justin Huang and So I Kwak"
date: 4-12-2020
site: bookdown::bookdown_site
---

# Libraries

```{r message=FALSE, warning=FALSE}
rm(list=ls())

# install.packages(c("tidyquant", "car", "cowplot", "wrapr", "Hmisc"))
library(tidyverse)
library(tidyquant)
library(car)
library(cowplot)
library(wrapr)
library(Hmisc)
```

# Guiding Question

Has the advent of COVID-19 had a significant impact on futures prices for Brent Crude Oil? If so, how has it been affected and where might it go in the foreseeable future?

# Primary data: COVID-19 Cases

### Load Global Data 

*uses absolute file paths, change for your system*
```{r}
dir = "~/Development/stat184/brent-covid-19/COVID-19/csse_covid_19_data/csse_covid_19_time_series"

Confirmed <- 
  read.csv(file.path(dir,
                     "time_series_covid19_confirmed_global.csv")) %>%
    pivot_longer(cols=c(starts_with("X")), names_to="date", values_to="confirmed")

Deaths <- 
  read.csv(file.path(dir,
                     "time_series_covid19_deaths_global.csv")) %>%
    pivot_longer(cols=c(starts_with("X")), names_to="date", values_to="deaths")

Recovered <- 
  read.csv(file.path(dir,
                     "time_series_covid19_recovered_global.csv")) %>%
    pivot_longer(cols=c(starts_with("X")), names_to="date", values_to="recovered")

# get all data in one data table, and make everything tidy
COVID19 <-
  Confirmed %>%
    left_join(Deaths %>% select(-c(Lat, Long)), 
              by=c('Province.State' = 'Province.State', 
                   'Country.Region' = 'Country.Region', 
                   'date'='date')) %>%
    left_join(Recovered %>% select(-c(Lat, Long)),
              by=c('Province.State' = 'Province.State', 
                   'Country.Region' = 'Country.Region', 
                   'date'='date')) %>%
    rename(c("province.state"="Province.State", 
             "country.region"="Country.Region", 
             "lat"="Lat", 
             "long"="Long")) %>%
    mutate(date = lubridate::mdy(gsub(pattern = "^X", replacement="", date)))
```

### Summary of Primary Data

```{r}
summary(COVID19)
tail(COVID19)
```

We have three categorical variables (province.state, country.region, date) and three quantitative variables that describe the number of people affected by the coronavirus (confirmed, deaths, recovered). Additionally, we have lat and long but I declined to include these as they're not extremely pertinent to the overall analysis.

### Analysis

* Data found from the JHU COVID-19 repo [here](https://github.com/CSSEGISandData/COVID-19); forked and included as a submodule in this repository
* Collected by the Johns Hopkins CSSE group, sourced by various health organizations arouond the world
* Data collection and posting started on January 22, 2020 and collected to keep the public informed about the rapidly-developing COVID-19 crisis
* Case represents the number of COVID-19 confirmed cases, deaths and recoveries for a given province in a given country on a given day. In other words, it's the confirmed cases, deaths and recoveries for every combination of a country's regions and dates. Notably, the numbers are cumulative. There are a total of 19,166 cases available.
* If we use some sort of map display, the latitude and longitude data will be useful. Otherwise, the region, country, confirmed cases, deaths, and recoveries are my primary interest for answering the guiding question.

# Secondary Data

### Load Brent Crude Oil
```{r}
Brent <- 
  tq_get("BZ=F", from=mdy("1.22.2020")) %>%
    select(-symbol) %>%
    # Get rid of NA rows where trading wasn't open
    drop_na()
```

Conveniently, this is already in tidy form, so there's a lot less data wrangling to initially do.

### Summary of Secondary Data

```{r}
tail(Brent)
summary(Brent)
```

We have one categorical variable (date) and the rest are quantitative variables (OHLC values, volume, adjusted close). You'll notice that there were a good amount of NA rows because trading wasn't available on that specific day, so we cleaned those up.

### Analysis

* Data found using the `tidyquant` package API which pulls from Yahoo Finance
* Yahoo Finance maintains the data
* Data collection starts from January 20, 2020 to match the COVID-19 data, and is maintained for investors and traders to keep track of financial changes
* One case represents the OHLC end-of-day Brent crude oil values for each day (notice that the days where trading is closed is ommitted). There are 62 cases available.
* Plan is to mostly pay attention to date, close and volume variables. We're going to ignore adjusted values just because no splits occur with futures so it should be fine.

# Preliminary Exploration

### COVID-19

#### Global Comparison

```{r fig.height=6, fig.width=12, warning=FALSE}
COVID19 %>%
  filter(country.region %in% c("US", "China", "Korea, South", "Italy")) %>%
  group_by(country.region, date) %>%
  summarise(confirmed=sum(confirmed)) %>%
  ggplot(aes(date, confirmed)) +
    geom_line(aes(color=country.region)) +
    scale_y_log10()+
    theme_minimal()
```
Log scale of coronavirus cases in China, Italy, South Korea and the United States. We can see that China and South Korea have long since plataued in new confirmed cases, while the US and Italy are still in the thick of it. Something very noteworthy is how cases in Italy and South Korea began rising at approximately the same date, yet South Korea has been able to control the situation much more effectively than Italy. Of course, there may be other factors involved. In any case, a lot of these numbers are beginning to taper off, which may be indicative of the effectiveness of social distancing.

#### Confirmed, Deaths, Recoveries

```{r fig.height=6, fig.width=12, warning=FALSE}
Global <- 
  COVID19 %>%
    group_by(date) %>%
    summarise(confirmed=sum(confirmed), deaths=sum(deaths), recovered=sum(recovered, na.rm=TRUE))

Global %>%
  pivot_longer(cols=c(confirmed, deaths, recovered), names_to="type", values_to="count") %>%
  ggplot(aes(date, count)) +
    geom_line(aes(color=type)) +
    scale_y_log10() +
    theme_minimal()
```

Log scale of confirmed cases and deaths from COVID-19 globally. We can see that they essentially follow the same trends, which makes sense. Recoveries actually seem to grow faster than deaths, which is expected but still heartening.

### Brent Crude Oil

```{r fig.height=6, fig.width=12, warning=FALSE}
Brent %>%
  ggplot(aes(date, close)) +
    geom_line() +
    geom_candlestick(aes(open=open, high=high, low=low, close=close)) +
    theme_tq()
```

We can see that Brent crude oil futures prices tanked starting in January. I wonder what happened? Hmm. 

Despite its hardships, it's also begun to rebound well since the beginning of April.

One key thing to note about the price of Brent crude is the price war between Russia and OPEC that first started on March 6, when Russia refused to cut production in conjunction with Saudia Arabia because of a sharp fall in demand because of COVID-19 and stay-at-home procedures. While an agreement between Russia and Saudi Arabia was reached on April 9 to cut production by 10 million barrels per day, the situation is still evolving because of a lack of commitment from Mexico (source: [Bloomberg](https://www.bloomberg.com/news/articles/2020-04-10/mexico-reaches-oil-output-cuts-agreement-with-opec-and-trump)).

```{r fig.height=6, fig.width=12, warning=FALSE}
Brent %>%
  ggplot(aes(date, adjusted)) +
    geom_line(color = "red") +
    geom_point(color = "red") +
    geom_vline(aes(xintercept=mdy("3-6-2020")), linetype=3) +
    geom_vline(aes(xintercept=mdy("3-31-2020")), linetype=3) +
    geom_vline(aes(xintercept=mdy("4-9-2020")), linetype=3) +
    xlim(mdy("3-1-2020"), mdy("4-9-2020")) +
    theme_tq()
```

In any case, the sharp fall in Brent futures prices in early March may be attributed to the Russia-OPEC feud, and the rebound in early April may be due to news about talks of slashing production by 20 million barrels per day. Note particularly that on April 9, the high value and close are significantly further apart because of a rise in prices from the agreement, and subsequent fall after hearing about Mexico and possibly that production was only cut by 10 million barrels per day, which some investors may deem to be too conservative. 

While this price war was certainly caused by a fall in demand due to COVID-19, it's important to keep this information in mind moving forward.

Oh, and we should probably know how much Brent fell at its worst.

```{r}
(min(Brent$close) - max(Brent$close)) / max(Brent$close)
```

Down -64%? Cool!

### COVID-19 and Brent Crude Oil

```{r fig.height=6, fig.width=12, warning=FALSE}
Global <-
  Global %>%
    mutate(confirmed_delta = (confirmed/lag(confirmed)-1)*100,
         deaths_delta = (deaths/lag(deaths)-1)*100,
         recovered_delta = (recovered/lag(recovered)-1)*100)

Brent <-
  Brent %>%
    mutate(close_delta = (close/lag(close)-1)*100)

Global %>%
  ggplot(aes(x=date)) +
    geom_area(aes(y=confirmed_delta), fill = "blue", alpha = 0.5) +
    geom_area(data = Brent, aes(date, close_delta), fill = "green", alpha = 0.5) +
    scale_y_log10() +
    ylab("percent change")
```

In this graph, it's hard to see any correlation between percent change in COVID-19 confirmed cases and Brent Crude Oil prices. Let's try again.

```{r fig.height=6, fig.width=12, warning=FALSE}
COVID19 %>%
  filter(country.region=="China") %>%
  group_by(date) %>%
  summarise(confirmed=sum(confirmed)) %>%
  mutate(confirmed_delta = (confirmed/lag(confirmed)-1)*100) %>%
  ggplot(aes(date)) +
    geom_area(aes(y=confirmed_delta), fill="blue", alpha = 0.5) +
    geom_area(data = Brent, aes(date, close_delta), fill="red", alpha = 0.5) +
    scale_y_log10() +
    ylab("percent change")
```

This one's interesting, because you can see more of a negative relationship between percent change in close prices and China's confirmed cases (scaled by log10). However, this might be a red herring because the actual percent change of China in March and April is close to 0, and we all know that log(0) is not pretty.

# EDA: Projections

When modeling populations and growth, the logistic growth model becomes very handy. In the following examples, we will first use Hubei, China as an example of how the logistic growth model matches up with our data, and then project values for Italy and US. We'll also see how the data matches with Brent.

```{r}
logistic_model = function(xvariable, yvariable, asym, df) {
  #xvar <- enquo(xvar)
  #yvar <- enquo(yvar)
  let(
    c(xvar=xvariable, yvar=yvariable),
    {
      start <- coef(lm(logit(yvar / 100) ~ xvar, data = df))
      
      nls_model <- nls(yvar ~ phi1 / (1 + exp(-(phi2 + phi3 * xvar))),
                       start = list(phi1 = asym, phi2 = start[[1]], phi3=start[[2]]), 
                       data = df)
      
      rss <- sum(residuals(nls_model)^2)
      tss <- sum((df$yvar - mean(df$yvar))^2)
      r2 <- abs(tss-rss)/tss # Calculation of r^2
    }
  )
      
  # Values for the logistic growth model
  phi1 <- coef(nls_model)[1]
  phi2 <- coef(nls_model)[2]
  phi3 <- coef(nls_model)[3]
  
  # Get x value range
  x <- c(min(df[xvariable]):(max(df[xvariable])+50))
      
  # Construct model return values and return the resulting data frame
  y <- phi1 / (1 + exp(-(phi2 + phi3 * x)))
  
  return(list(data.frame(x, y), r2))
}
```

The function above is inspired by the logistic growth model example by [Brian Cheng](https://bscheng.com/2014/05/07/modeling-logistic-growth-data-in-r/). 

## Hubei, China

```{r}
brent_graph <- 
  Brent %>%
    mutate(datediff = as.numeric(date - mdy("1-22-2020"))) %>%
    ggplot(aes(datediff, close)) +
      geom_line() +
      labs(x='Days Since 1-22-2020', y='Brent Futures Close Price') +
      theme_minimal()
```

```{r fig.height=6, fig.width=12, warning=FALSE}
Hubei <-
  COVID19 %>%
    filter(province.state=="Hubei") %>%
    mutate(datediff = as.numeric(date - mdy("1-22-2020")))

# See function definition above.
hubei_logit <- logistic_model("datediff", "confirmed", 100000, Hubei)

hubei_graph <-
  Hubei %>%
    ggplot(aes(datediff, confirmed))+
      geom_point(color='blue') +
      labs(x='Days Since 1-22-2020', y='Confirmed Cases')+
      geom_line(data=hubei_logit[[1]], aes(x, y)) +
      geom_vline(aes(xintercept=0), linetype = 3) +
      geom_vline(aes(xintercept=40), linetype = 3) +
      xlim(0, 60) +
      theme_minimal()

plot_grid(hubei_graph, 
          brent_graph +
          geom_vline(aes(xintercept=0), linetype = 3) +
          geom_vline(aes(xintercept=40), linetype = 3) +
          xlim(0, 60))
```

This Hubei examples exists to show that the logistic growth model seems to fit COVID-19 confirmed cases over time 
well&mdash;visually, at least. Let's check that it actually works.

```{r}
hubei_logit[[2]]
```

That's a beautiful r-squared value. 

Interestingly, there is little effect on the Brent Crude price, since in the range where the cases grew the most, there was almost no net change in Brent Crude up to day 30 (since the prices rebounded) and then began to dip as cases in China began to stabilize. Yes, there is a significant dip, but that came *after* the majority of the growth in Chinese cases. So what *did* affect the oil price?

## Italy

```{r fig.height=6, fig.width=12, warning=FALSE}
Italy <-
  COVID19 %>%
    filter(country.region=="Italy") %>%
    mutate(datediff = as.numeric(date - mdy("1-22-2020")))

italy_logit <- logistic_model("datediff", "confirmed", 1000000, Italy)

italy_graph <-
  Italy %>%
    ggplot(aes(datediff, confirmed))+
      geom_point(color='blue') +
      labs(x='Days Since 1-22-2020', y='Confirmed Cases') +
      geom_line(data=italy_logit[[1]], aes(x, y)) +
      geom_vline(aes(xintercept=37.5), linetype=3) +
      geom_vline(aes(xintercept=90), linetype=3) +
      xlim(0, 120) +
      theme_minimal()

plot_grid(italy_graph, brent_graph + geom_vline(aes(xintercept=37.5), linetype=3))

italy_logit[[2]]
```

On a more fundamental level, Italy was the first indication of how much of an impact the coronavirus would have on the Western world, which sent a lot of markets into a frenzy. The dashed line indicates the day that cases began picking up, and we also observe that prices began dropping rapidly. Once again, we make the clarification that the Russia-OPEC price war drove those prices to the ground, but Italy was the sign that things were going to be bad.

Back to real statistics now. The r-squared value is indeed high, but some may be dissuaded by the divergence of the model's projections from the last recorded value of confirmed cases in Italy. From there, we turn to the logarithmically-scaled values to double-check.

```{r fig.height=6, fig.width=12, warning=FALSE}
Italy %>%
  ggplot(aes(datediff, confirmed)) +
    geom_line(color="blue") +
    scale_y_log10() +
    geom_line(data=italy_logit[[1]], aes(x, y), color="red") +
    theme_minimal()
```

And the tail of the recorded cases almost perfectly matches the curve of the logistic model projections, which is a good sign.

## US

```{r fig.height=6, fig.width=12, warning=FALSE}
America <-
  COVID19 %>%
    filter(country.region=="US") %>%
    mutate(datediff = as.numeric(date - mdy("1-22-2020")))

us_logit <- logistic_model("datediff", "confirmed", 150000000, America)

us_graph <-
  America %>%
    ggplot(aes(datediff, confirmed))+
      geom_point(color='blue') +
      labs(x='Days Since 1-22-2020', y='Confirmed Cases') +
      geom_line(data=us_logit[[1]], aes(x, y)) +
      geom_vline(aes(xintercept=50), linetype=3) +
      geom_vline(aes(xintercept=72), color = "lightblue") +
      geom_vline(aes(xintercept=95), linetype=3) +
      xlim(40, 120) +
      theme_minimal()

plot_grid(us_graph, brent_graph + geom_vline(aes(xintercept=72), color="lightblue") + xlim(40, 80))

us_logit[[2]]
```

The r-squared value is almost too perfect. 

Based on this model, we'll expect a plateau to come around the 95th day, or...
```{r}
mdy("1-22-2020") + 95
```

So essentially, the end of April. 

It's important to note that this model gives a very low estimate of the carrying capacity in the US, which seems to plateau a little above 600000 cases. We've also passed the model's inflection point (around 72 days into data recording). Funnily enough, that inflection came the day Brent rebounded.

I can't say exactly how great that projection is because it assumes:

* A relatively low asymptote
* All reported cases are accurate of all coronavirus cases in the US
* The status quo will not change (in this case, social distancing and stay-at-home orders)

Moving on, lookin at the comparison betweeen America and Brent Crude futures prices, while the downfall of Brent did occur during the rise of cases in America, we still keep in mind that the fall was driven a lot by the price war in Saudi Arabia and Russia. That being said, the skyrocketing cases in America most likely did play a role in the fall of Brent crude oil because of the financial superpower in the country.

## Global

The logistic model is a little difficult to use in this case because of nonlinear parameter estimation, so we'll use the classic smoother instead.

```{r fig.height=6, fig.width=12, warning=FALSE}
global_graph <-
  Global %>%
    mutate(datediff = as.numeric(date - mdy("1-22-2020"))) %>%
    pivot_longer(cols=c(confirmed, deaths, recovered), names_to="type", values_to="count") %>%
    ggplot(aes(date, count, color=type))+
      geom_point() +
      geom_smooth(span = 0.5) +
      labs(x='Days Since 1-22-2020', y='Confirmed Cases') +
      theme_minimal()

plot_grid(global_graph, brent_graph)
```

A normal stat smooth seems to overfit the data. The confidence bands are "missing" because they're actually so tiny that you can't see them&mdash;another bad sign of overfitting the data.

```{r fig.height=6, fig.width=12, warning=FALSE}
Global %>%
  mutate(datediff = as.numeric(date - mdy("1-22-2020"))) %>%
  pivot_longer(cols=c(confirmed, deaths, recovered), names_to="type", values_to="count") %>%
  ggplot(aes(date, count, color=type))+
    geom_point() +
    # This is an insane confidence level, yet the bands are just barely visible.
    geom_smooth(span = 0.5, level=0.999999999999999) +
    labs(x='Days Since 1-22-2020', y='Confirmed Cases') +
    theme_minimal() +
    scale_y_log10()
```

Even on a logarithmic scale, the problem is still there.

# Conclusion

Recall that our original guiding question revolves around the analysis of effects of COVID-19 cases on Brent crude oil futures prices.

We've examined historical data from COVID-19 recorded cases and Brent crude oil futures prices. From our analysis, we can see that Brent fell 64% at its worst as COVID-19 confirmed cases around the world climbed higher and higher. Fundamentally, this can be explained by a severe demand crash in oil as people stopped traveling and stayed at home instead. We also know that the Russia-OPEC price war severely affected the confidence in Brent as they began ramping up production instead of cutting it&mdash;although that problem will be resolved soon, by the looks of it.

Statistically, we want to project how COVID-19 will affect the world in the near future. To that end, we used the logistic growth model to model our confirmed cases. First, we showed its fit on Hubei data, which had an extremely high r-squared value, and oil prices began showing signs of dropping. Then we have Italy, the first serious foray of COVID-19 into the Western world, and the United States. In the period where Italian cases were beginning to severely rise, oil prices began to drop rapidly, and the trend continued in almost a staircase pattern once cases in the United States began rising rapidly.

However, there's going to be an end to all of the cases. In fact, Italy seems to be reaching its peak extremely, extremely soon, while America will follow in late April. Once those numbers begin stabilizing (approach the asymptote of the logistic model's asymptote), we may see an increase in Brent crude oil futures prices once again as they stabilize.

It's important to note the limitations of the logistic growth model used in projections. The most important one is the assumption of stasis in conditions used in coronavirus prevention&mdash;namely, social distancing. If we were to resume normal conditions prematurely, the logistic growth model used here would no longer apply, since the carrying capacity would suddenly increase as more virus exposure becomes possible. 

The model also heavily relies on accurate data reporting by global health organizations, as they are the ones providing the data (compiled by Johns Hopkins). This doesn't necessarily mean that they are purposefully lowering the numbers, but that there are other unreported cases by people who do not go to the appropriate healthcare infrastructure, or simply not classified as infected yet.

In any case, we conclude that the rise of the coronavirus since January has severely affected Brent crude oil futures prices, mixed in with current events and investor confidence, but according to our projections, we may see Brent crude stabilize towards the end of the month as COVID-19 confirmed cases begin to plateau, so long as we maintain the current prevention measures.
